---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidyheatmaps

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/jbengler/tidyheatmaps.svg?branch=master)](https://travis-ci.com/jbengler/tidyheatmaps)
[![Codecov test coverage](https://codecov.io/gh/jbengler/tidyheatmaps/branch/master/graph/badge.svg)](https://codecov.io/gh/jbengler/tidyheatmaps?branch=master)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of `tidyheatmap` is to provide a tidyverse-style interface to the powerful heatmap package [pheatmap](https://github.com/raivokolde/pheatmap) by [\@raivokolde](https://github.com/raivokolde). This enables the convenient generation of complex heatmaps from tidy data with minimal code.

## Installation

You can install `tidyheatmap` from GitHub with:

```{r eval=FALSE}
# install.packages("devtools")

devtools::install_github("jbengler/tidyheatmaps")
```

## Example

Given a tidy data frame of gene expression data like `data_exprs`, you can easily generate a customized heatmap.

```{r fig.retina=2}
library(tidyheatmaps)

tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             scale = "row",
             annotation_col = c(sample_type, condition, group),
             annotation_row = c(is_immune_gene, direction),
             gaps_row = direction,
             gaps_col = group
)
```

## Documentation

https://jbengler.github.io/tidyheatmaps/


