---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidyheatmaps <img src="man/figures/logo.svg" align="right" height="139" alt="" />

<!-- badges: start -->
<!-- badges: end -->

tidyheatmaps simplifies the generation of publication-ready heatmaps from tidy data. By offering an interface to the powerful [pheatmap](https://github.com/raivokolde/pheatmap) package, it allows for the effortless creation of intricate heatmaps with minimal code.

## Installation

``` r
# install.packages("devtools")
devtools::install_github("jbengler/tidyheatmaps")
```

## Usage

Given a tidy data frame of gene expression data like `data_exprs`, you can easily generate a customized heatmap. The full documentation can be found [here](https://jbengler.github.io/tidyheatmaps/).

```{r fig.retina=2}
library(tidyheatmaps)

tidyheatmap(data_exprs,
            rows = external_gene_name,
            columns = sample,
            values = expression,
            scale = "row",
            annotation_col = c(sample_type, condition, group),
            annotation_row = c(is_immune_gene, direction),
            gaps_row = direction,
            gaps_col = group
)
```


## Acknowledgements

tidyplots relies on a number of fantastic packages that do all the heavy lifting behind the scenes. These include 
    dplyr,
    pheatmap,
    rlang,
    grDevices,
    tidyr,
    tibble and
    RColorBrewer.

